---
title: "NFL Losers Pool 2021 Optimal Pick Algorithm"
author: "Jordan Hutchings"
date: "08/11/2021"
output: html_document
---

```{r, echo = FALSE}
# setup ----
pacman::p_load(dplyr, ggplot2, kableExtra, data.table)



```

I've joined my buddies annual NFL Losers Pool, and while I spend much less time 
keeping up with the NFL than other competitors in the pool, I have a secret 
strategy in mind. Now that the pool is over, I am able to publicize the approach 
I took to maximizing my likelihood of success in the pool. 

For those who haven't heard of a losers pool before, the rules are pretty straight 
forward. Each entrant picks one team a week to lose. If your pick wins, you are 
eliminated from the pool. The last entrant left in the pool collects the total 
pot of entry fees. The one catch, once a team is picked, you cannot pick that 
team again. This is where the strategy comes into play, by optimizing when to 
pick each team, one can reduce their likelihood of losing from the pool. 

While it would have been a great exercise to forecast the outcomes of the NFL season, 
I've instead choose to leverage the NFL game projections made by 
[FiveThirtyEight](https://projects.fivethirtyeight.com/2021-nfl-predictions/games/). 
These projections are based off their quarterback-adjusted ELO forecasts. 

### ELO Forecasts

ELO is a fantastic and widely used method of ranking two competing teams. 
Commonly used in chess, your ELO score is a representation of your skill based 
on your performance against other ELO ranked players. There is a great video 
describing the ELO process made by James Grime on his 
[youtube channel](https://www.youtube.com/watch?v=AsYfbmp0To0&ab_channel=singingbanana). 
All you really need to know about ELO is that we can derive the probability 
a given team A beats the other team B as: 

$$Pr(\text{A Wins}) = \frac{1}{1+10^{(ELO_B -ELO_A)/400}} $$
The FiveThirtyEight team has compiled the ELO rankings over time for each NFL 
team based on their past results, and uses these scores - with some modifications 
such as, home advantage, and skill of the quarterback - to forecast the average 
win probabilities for each game of the season. 

We can verify the performance of using ELO to forecast game outcomes by plotting 
the frequency of game outcomes based on their ELO projections. If ELO were a good 
forecasting variable, we would expect to see a 45 degree line between projected and 
observed outcomes. 

```{r}
read.csv("https://projects.fivethirtyeight.com/nfl-api/nfl_elo.csv") %>%
  mutate(loser_score = ifelse(qbelo_prob1 > qbelo_prob2, score2, score1), 
         winner_score = ifelse(qbelo_prob1 > qbelo_prob2, score1, score2), 
         p_win_winner = ifelse(qbelo_prob1 > qbelo_prob2, qbelo_prob1, qbelo_prob2), 
         win = as.numeric(winner_score > loser_score), 
         win_bucket = round(p_win_winner, 2)) %>%
  filter(!is.na(winner_score)) %>%
  group_by(win_bucket) %>%
  summarise(p = mean(win)) %>%
  ggplot(aes(x = win_bucket, y = p)) + 
  geom_point() + 
  geom_smooth(alpha = 0.7, linetype = "dashed", se = F, method = "loess", formula = "y ~ x") + 
  geom_abline(slope = 1) + 
  annotate("text", x = 0.5, y = 0.49, label = "y = x", hjust = 0) + 
  annotate("text", x = 0.5, y = 0.6, label = "Line of \nbest fit", hjust = 0, color = "blue") + 
  labs(title = "The historical oucome observation lines up well with the projected outcomes", 
       x = "Forecasted win rate", 
       y = "Observed win rate") + 
  theme_bw(12)
```


```{r, echo = FALSE}
# Helper Functions ----
read_data = function(path, past_picks, all_weeks = FALSE){
  dt = data.table::fread(path)
  week1 <- as.Date("2021-09-09")
  
  # calculate week, get proj loser and prob of loss. 
  
  dt[, loser:=ifelse(qbelo_prob1 > qbelo_prob2, team2, team1)]
  dt[, winner:=ifelse(qbelo_prob1 > qbelo_prob2, team1, team2)]
  dt[, p_win:=ifelse(qbelo_prob1 > qbelo_prob2, qbelo_prob2, qbelo_prob1)]
  dt[, p_win_winner:=ifelse(qbelo_prob1 > qbelo_prob2, qbelo_prob1, qbelo_prob2)]
  dt[, week:=floor(as.numeric(difftime(date, week1, units="days")) / 7) + 1]
  dt[, loser_score:=ifelse(qbelo_prob1 > qbelo_prob2, score2, score1)]
  dt[, winner_score:=ifelse(qbelo_prob1 > qbelo_prob2, score1, score2)]
  dt = dt[, upset:=ifelse(loser_score > winner_score, TRUE, FALSE)]
  
  dt = dt[, .(week, loser, winner, p_win, p_win_winner, upset)]
  
  
  
  # if(!all_weeks){
  #   dt = dt[week %in% time_period]
  #   dt = dt[!loser %in% past_picks]
  # }
  
  return(dt)
}
```

There are 18 weeks and 32 teams to pick from in the 2021 season. From the 
heat map below, we see that Tom Brady and the Tampa Bay Buccaneers are favourited 
the most by the ELO rankings, and the Houston Texans are the forecasted worst team 
in the league. It is also worth noting, Houston has their lowest win rate of the 
season in Week 7, roughly the same rate as Detroit.

```{r}
fivethirtyeight = "https://projects.fivethirtyeight.com/nfl-api/nfl_elo_latest.csv"
plot_data = read_data(fivethirtyeight, all_weeks = TRUE)

plot_data %>% 
  tidyr::pivot_longer(cols = c("loser", "winner"), values_to = "team") %>%
  mutate(p = ifelse(name == "loser", p_win, p_win_winner)) %>%
  select(-c("p_win", "p_win_winner")) %>%
  group_by(team) %>%
  mutate(win_avg = mean(p)) %>%
  ungroup() %>%
  mutate(team = reorder(team, win_avg)) %>% # reorder teams by avg win
  ggplot(aes(week, team, fill=p)) + 
  geom_tile() + 
  scale_fill_viridis_c("Prob. Win") + 
  theme_bw() + 
  labs(title = "Weekly Win Probabilities Heatmap by team", 
       x = "Week Number", 
       y = "Team Name") + 
  scale_x_continuous(expand = c(0, 0))
```

### Determining the optimal pick profile

My approach for the contest was to pick teams based on their opportunity cost of 
not being selected, or in other words, the percentage differential towards the 
next best team to pick in a given week. Additionally, since lasting week to week 
is the most important aspect of the losers pool, I discount the opportunity cost 
so that future weeks are less important than upcoming weeks. 

I also compare the results from my opportunity cost algorithm against a naive 
algorithm that picks the lowest probability team to win each week. 

Another caveat of the contest is entrants are allowed to re-enter up until week 
3. Due to the scarcity of teams in the pool, I run the model beginning at Week 3, 
and then pick from the remainder of teams in Weeks 1 and 2. 

Last year, the pool lasted only until Week 10. For this reason, I also do not 
optimize my picks beyond this week. The pool is less about lasting to Week 18 as 
it is about outlasting the other competitors. 

```{r}
# Picking functions ----
join_picks <- function(past_weeks, past_picks){
  picks <- setDT(data.frame(week = past_weeks, loser = past_picks))
  picks <- merge(picks, dt, on = week)
  picks <- picks[, .(week, loser, p_win)]
  
  return(picks)
}

by_oc <- function(dt, past_picks, past_weeks, start_week = 3, total_weeks = 10, beta = 1){
  
  for(i in start_week:total_weeks){
    
    pick <- dt[week %in% c(start_week:total_weeks)]
    pick <- pick[!week %in% past_weeks & !loser %in% past_picks, ]
    pick <- pick[order(week, p_win)]
    pick[, oc:= shift(p_win, 1, type="lead") - p_win, by = week]
    pick[, oc:= oc * beta^(week - start_week)]
    pick <- pick[, .SD[1], week]
    pick <- pick[order(-oc)]
    pick <- pick[, .SD[1]]
    
    past_weeks <- append(past_weeks, pick$week)
    past_picks <- append(past_picks, pick$loser)
    
  }
  
  picks <- join_picks(past_weeks, past_picks)
  
  return(picks)
}

by_week <- function(dt, past_picks, past_weeks, start_week = 3, total_weeks = 10, beta = 1){
  # loop through filling in each week
  for(i in start_week:total_weeks){
    pick <- dt[!loser %in% past_picks & week == i]
    pick[, value:= p_win * beta^(week - start_week)]
    pick <- pick[order(week, value)]
    pick <- pick[, .SD[1]]
    
    past_weeks <- append(past_weeks, pick$week)
    past_picks <- append(past_picks, pick$loser)
    
  }
  
  picks <- join_picks(past_weeks, past_picks)
  
  return(picks)
}

cumprob <- function(picks, inc_weeks = FALSE){
  p <- purrr::accumulate((1-picks$p_win), function(x, y)  x * y)
  weeks <- picks$Week
  if(inc_weeks){ out <- data.frame(p, weeks) } 
  else { out <- data.frame(p) }
  
  out
}
```


```{r, echo = FALSE, message = FALSE}
dt = read_data("https://projects.fivethirtyeight.com/nfl-api/nfl_elo_latest.csv", past_picks = c())
past_weeks = c(1:2) # skip weeks 1 and 2
past_picks = c("DAL", "TEN") # My first picks after optimizing

pick = by_oc(dt, past_picks, past_weeks, beta = 0.9)
naive = by_week(dt, past_picks, past_weeks)

tbl = pick %>% 
  left_join(naive, by = "week", suffix = c("_oc", "_week")) %>%
  mutate(delta = p_win_oc - p_win_week)

names(tbl) = c("Week", rep(c("Team", "ProbWin"), 2), "Diff")

kbl(tbl, digits=3, caption = "Optimal Picks per Week") %>%
  kable_classic(full_width=F) %>%
  add_header_above(c(" " = 1, "Oppertunity Cost" = 2, "Lowest Per Week" = 2, " " = 1))
```

Given the Boolean nature of our predictions, we can compute the likelihood 
of reaching a given week based on our pick paths. Surprisingly, the By Week 
model outpreforms the Oppertunity cost model up until Week 13. 

```{r, echo = FALSE, message = FALSE}
models = list(pick[3:10, ], naive[3:10, ])
models = lapply(models, cumprob) %>% 
  bind_cols()
models$week = 3:10
names(models) = c("Oppertunity Cost", "By Week", "Week")
models %>%
  tidyr::pivot_longer(cols = -Week, names_to = "Model", values_to = "p") %>%
  ggplot(aes(x = Week, y = p, color = Model)) +
  geom_line() + 
  geom_point(alpha = 0.8) + 
  scale_x_continuous(expand = c(0, 0), limits = c(3, 10)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) + 
  scale_color_viridis_d() + 
  labs(title = "Probability of reaching a given week, for the first 10 weeks",
       x = "Week number", 
       y = "Pr(W < w)") + 
  theme_bw(12)
```

As we can see above, we slightly out preform simply picking the most likely team 
to win each week. Where the Opportunity Cost model preforms is in later weeks when 
picks begin to thin out. 

### Season Results

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ownership <- "https://docs.google.com/spreadsheets/d/1sajv1HXDqzjG2bXwx927MTP_TUKRWsjVSvKAwnaWZcs/edit?usp=sharing"
own <- googlesheets4::read_sheet(ownership)
own <- tidyr::pivot_longer(own, cols = contains("Week"), names_to = "Week", values_to = "Pick")

#TODO: update to show player is out
own = own %>% 
  mutate(Week = as.numeric(gsub("Week ", "", Week)), 
         OUT = ifelse(Pick == "OUT", TRUE, FALSE)) %>%
  group_by(Player, Week) %>%
  left_join(dt, by = c("Week" = "week", "Pick" = "loser")) %>%
  select("Week", "Pick", "p_win", "Player", "OUT", "upset")
  

ggplot(subset(plot_data, week < 11), aes(x = week, y = p_win)) + 
  geom_point(size = 2, alpha = 0.6, color = "#440154FF") + 
  geom_line(data = own, mapping = aes(x = Week, y = p_win, group = Player), alpha = 0.3) + 
  geom_point(data = own, mapping = aes(x = Week, y = p_win, color = upset), size = 2, alpha = 0.9, stroke = 1) + 
  coord_flip() + 
  xlim(1, 10) + 
  # annotate("text", x = 1 + 0.05, y = 0, label="forecast \nhistorical", hjust = 0, color = "darkgrey") + 
  labs(title = "Choice set path of all contestants, with upsets", 
       y = "Win Probability", 
       x = "Week Number") + 
  theme_bw(12) +
  theme(legend.position="none") + 
  scale_color_manual(values=c("#440154FF", "#FDE725FF"))


```

The season only lasted to Week 9, with three players splitting the pot. 
